\textbf{Machine learning classification using self-reference-based OTU clustering}

### 2023-04-13 submission/manuscript.tex
first round revisions after pat comments

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Streamlined implementation of a machine learning model to
classify screen relevant neoplasia using reference-based OTU
clustering}}
\author{}
\date{\vspace{-2.5em}}

@@

Machine learning classification of disease using the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The abundance
of each OTU is then used to train a classification model. The standard
de novo approach to clustering sequences into OTUs leverages the
similarity of the sequences to each other rather than to a reference
database. However, such an approach depends on the sequences in the
dataset and therefore OTU assignments can change if new data are added.
This lack of stability complicates classification because in order to
use the model to classify additional samples, the new sequences must be
reclustered with the old data and the model must be retrained with the
new OTU assignments. The new reference-based clustering algorithm,
OptiFit, addresses this issue by fitting new sequences into existing
OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
whether this method for fitting new sequence data into existing OTUs
will impact the performance of classification models. We used OptiFit to
cluster additional data into existing OTU clusters and then evaluated
model performance in classifying a dataset containing samples from
patients with and without colonic screen relevant neoplasia (SRN). We
compared the performance of this model to the standard procedure of de
novo clustering all the data together. We found that both approaches
performed equally well in classifying SRNs. OptiFit can streamline the
process of classifying new samples by avoiding the need to retrain
models using reclustered sequences.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, such as colorectal cancer
(\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-zackular2014}{2}). Taxonomic composition of
microbial communities can be assessed using amplicon sequencing of the
16S rRNA gene, which is the input to classification models. Analysis of
16S rRNA gene sequence data generally relies on similarity-based
clustering of sequences into operational taxonomic units (OTUs). The
process of OTU clustering can either be reference-based or \emph{de
novo}. The quality of OTUs generated with reference-based clustering is
generally poor compared to those generated with \emph{de novo}
clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
novo} clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared two approaches, one using all the data to generate OTU
clusters, and the other generating \emph{de novo} OTU clusters with a
portion of the data and then fitting the remaining sequence data to the
existing OTUs using OptiFit. In the first approach, all the 16S rRNA
sequence data was clustered into OTUs with the OptiClust algorithm in
mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
abundance data was then split into training and testing sets, where the
training set was used to tune hyperparameters and ultimately train and
select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). However, this methodology requires
one to regenerate the OTU clusters and retrain the model to classify
additional samples. The OptiFit algorithm
(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
enabling new sequences to be clustered into existing OTUs. The OptiFit
workflow is similar to the OptiClust workflow, where the data was
clustered into OTUs and used to tune hyperparameters and ultimately
train the model. Then, we used OptiFit to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1B). To test how the model
performance compared between these two approaches, we used a publicly
available dataset of 16S rRNA gene sequences from stool samples of
healthy subjects (n = 261) as well as subjects with SRN consisting of
advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{1}). The dataset was randomly split
into an 80\% train set and 20\% test set. For the standard OptiClust
workflow, the training and test sets were \emph{de novo} clustered
together into OTUs, then the resulting abundance table was split into
the training and testing set. For the OptiFit workflow, the train set
was clustered \emph{de novo} into OTUs, and the remaining test set was
fit to the OTU clusters using the OptiFit algorithm. For both workflows,
the abundance table of the train set was used to tune hyperparameters
and train a random forest model to classify SRN. The test set was
classified as either control or SRN using the trained models. To account
for variation, the dataset was randomly split 100 times and the process
repeated for each of the 100 data splits. By comparing the model
performance of classifying the samples in the test dataset between the
OptiFit and OptiClust algorithms, we quantified the impact of using
OptiFit on model classification performance.

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
(\protect\hyperlink{ref-westcott2017}{5}). MCC scores range between
negative one and one. A score of negative one means none of the
sequences in an OTU are within the similarity threshold and any
sequences within the similarity threshold are not in an OTU together. An
MCC score of zero means the sequences are randomly clustered. An MCC
score of 1 means all sequences in an OTU are within the similarity
threshold and all sequence pairs within the similarity threshold are in
the same OTU. To ensure that OptiFit is appropriately integrating new
sequence data into the existing OTUs, we expected the MCC scores
produced by the OptiClust and OptiFit workflows to be similar. Since the
data was only clustered once in the OptiClust workflow there was only
one MCC score (MCC = 0.884) while the OptiFit workflow produced an MCC
score for the OTU clusters from each data split (average MCC = 0.879,
standard deviation = 0.002). As expected, the MCC scores were similar
between the two workflows. Another metric we examined for the OptiFit
workflow was the fraction of sequences from the test set that mapped to
the reference OTUs. Any sequences that did not map to reference OTUs
were eliminated therefore, if a high percentage of reads did not map we
might expect this loss of data to negatively impact classification
performance. We found that loss of data was not an issue since on
average 99.9\% (standard deviation = 0.004) of sequences in the test set
mapped to the reference OTUs. These results indicate that OptiFit
performed as well as OptiClust when integrating new sequences into the
existing OTUs.

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test dataset. To evaluate model performance, we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the OptiClust and
OptiFit approaches (p-value = 0.13; Figure 2A). The trained model was
then applied to the test data classifying samples as either control or
SRN. The performance on the test data was equivalent between the
OptiClust and OptiFit approaches (p-value = 0.38; Figures 2B and 2C)
indicating that new data could be fit to existing OTU clusters using
OptiFit without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
However, we demonstrated OptiFit can be used to fit new sequence data
into existing OTU clusters and performs equally well in predicting SRN
compared to \emph{de novo} clustering all the sequence data together.
The ability to integrate data from new samples into existing OTUs
enables the implementation of a single machine learning model. This is
important for model implementation because not all of the data needs to
be available or known at the time of model generation. Since results may
depend on the amount of data in the training set, further analysis is
needed to determine the number of samples that are necessary in training
to build a robust model capable of classifying diverse samples. A robust
machine learning model could be implemented as part of a non-invasive
and low-cost aid in diagnosing SRN and other diseases.

@@

\textbf{\emph{Dataset.}} Raw 16S rRNA gene sequence data isolated from
human stool samples was downloaded from NCBI Sequence Read Archive
(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
samples from a total of 490 subjects. For this analysis, samples from
subjects identified in the metadata as normal, high risk normal, or
adenoma were categorized as ``normal'', while samples from subjects
identified as advanced adenoma or carcinoma were categorized as ``screen
relevant neoplasia'' (SRN). The resulting dataset consisted of 261
normal samples and 229 SRN samples.

\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
and reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1).

The data was processed through two workflows. In the control workflow,
all the data was clustered together with OptiClust
(\protect\hyperlink{ref-westcott2017}{5}) to generate OTUs and the
resulting abundance tables were split into the training and testing
sets. In the experimental workflow, the preprocessed data was split into
the training and testing sets. The training set was clustered into OTUs
using OptiClust, then the test set was fit to the OTUs of the training
set using the OptiFit algorithm
(\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit algorithm was run
with method open so that any sequences that did not map to the existing
OTU clusters would form new OTUs. For both pathways, the shared files
were sub-sampled to 10,000 reads per sample.

\textbf{\emph{Machine Learning.}} A random forest model was trained with
the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{\emph{Code Availability.}} The analysis workflow was implemented
in Snakemake (\protect\hyperlink{ref-koster2012}{11}) . Scripts for
analysis were written in R (\protect\hyperlink{ref-R2020}{12}) and GNU
bash (\protect\hyperlink{ref-GNUbash}{13}). The software used includes
mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{14}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{15}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{16}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{17}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{18}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

@@

\includegraphics{../exploratory/figures/figure1_a.pdf}

\includegraphics{../exploratory/figures/figure1_b.pdf}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{helvet}
\renewcommand*\familydefault{\sfdefault}
\usepackage{setspace}
\doublespacing
\usepackage[left]{lineno}
\usepackage{multirow}
\usepackage[none]{hyphenat}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Machine learning classification using self-reference-based OTU clustering},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\textbf{Machine learning classification using
self-reference-based OTU clustering}}
\author{}
\date{\vspace{-2.5em}}

@@

Machine learning classification using the gut microbiome relies on
assigning 16S rRNA gene sequences into operational taxonomic units
(OTUs) to quantify microbial composition. OTU abundances are then used
to train a classification model that can be applied to classify new
samples. The standard approaches to clustering sequences include
reference-based and \emph{de novo} clustering. Reference-based
clustering requires a well-curated reference database that may not exist
for all systems. \emph{De novo} clustering tends to produce higher
quality OTU assignments than reference-based, but clusters depend on the
sequences in the dataset and therefore OTU assignments will change when
new samples are sequenced. This lack of stability complicates machine
learning classification since new sequences must be reclustered with the
old data and the model retrained with the new OTU assignments. The
OptiFit algorithm addresses these issues by fitting new sequences into
existing OTUs. While OptiFit produces high quality OTU clusters, it is
unclear whether this method for fitting new sequence data into existing
OTUs will impact the performance of classification models trained with
the older data. We used OptiFit to cluster sequences into existing OTUs
and evaluated model performance in classifying a dataset containing
samples from patients with and without colonic screen relevant neoplasia
(SRN). We compared the performance of this model to standard methods
including \emph{de novo} and database-reference-based clustering. We
found that using OptiFit performed as well or better in classifying
SRNs. OptiFit can streamline the process of classifying new samples by
avoiding the need to retrain models using reclustered sequences.

@@

There is increasing evidence for an association between the composition
of the gut microbiome and a variety of diseases, such as crohn's disease
and colorectal cancer. The signatures that associate with disease have
shown potential to aid in diagnosis of disease throught gut microbiome
profiling and machine learning. Taxonomic composition of microbial
communities can be assessed using amplicon sequencing of the 16S rRNA
gene, which is the input to classification models. Analysis of 16S rRNA
gene sequence data generally relies on assigning sequences into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or \emph{de novo}. The quality of OTUs
generated with reference-based clustering is generally poor compared to
those generated with \emph{de novo} clustering
(\protect\hyperlink{ref-westcott2015}{1}). While \emph{de novo}
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the sequences within the dataset and the addition of
new data has the potential to redefine OTU cluster composition. The
unstable nature of \emph{de novo} OTU clustering complicates deployment
of machine learning models since integration of additional data requires
reclustering all the data and retraining the model. The ability to
integrate new data into a validated model without reclustering and
retraining could allow for the application of a single model that can
continually classify new data. Recently, Sovacool \emph{et al.}
introduced OptiFit, a method for fitting new sequence data into existing
OTUs (\protect\hyperlink{ref-sovacool2022}{2}). While OptiFit can
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification
performance. Here, we tested the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of classifying
disease based on gut microbiome composition.

We compared the ability of several approaches for assigning 16S rRNA
gene seequences to OTUs including, \emph{de novo},
database-reference-based, and self-reference-based with OptiFit. To test
how the model performance compared between these approaches, we used a
publicly available dataset of 16S rRNA gene sequences from stool samples
of healthy subjects (n = 226) as well as subjects with screen-relevant
neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
(\protect\hyperlink{ref-baxter2016}{3}). For the \emph{de novo}
workflows, all the 16S rRNA sequence data was clustered into OTUs. The
OTU clustering was conducted using two common algorithms: 1) the
OptiClust algorithm in mothur 1 and 2) the VSEARCH algorithm used in
QIIME2 (\protect\hyperlink{ref-rognes2016}{4},
\protect\hyperlink{ref-bolyen2019}{5}). For both algorithms, the
resulting abundance data was then split into training and testing sets,
where the training set was used to tune hyperparameters and ultimately
train and select the model. The model was applied to the testing set and
performance evaluated (Figure 1A). We also conducted reference-based OTU
clustering using OptiFit to fit the sequence data into OTUs based on the
greengenes reference database. To compare with another commonly used
method, we also used the VSEARCH algorithm to fit the sequence data to
the greengenes reference (Figure 1B). In the OptiFit self-reference
workflow, the data was split into a training and a testing set. The
training set was clustered into OTUs and used to train a classification
model. The OptiFit algorithm was used to fit sequence data of samples
not part of the original dataset into the existing OTUs, and used the
same model to classify the samples (Figure 1C). For each of the
workflows the process was repeated for 100 random splits of the data to
account for variation caused by the choice of the random number
generator seed.

We first examined the quality of the resulting OTU clusters from each
method using the Matthews correlation coefficient (MCC). MCC is a metric
used to measure OTU cluster quality based on the similarity of all pairs
of sequences and whether they are appropriately clustered or not
(\protect\hyperlink{ref-westcott2015}{1}). MCC scores range between
negative one and one, and measure how well clustering assignment
correlates with the distance between sequences. To ensure that OptiFit
appropriately integrated new sequence data into the existing OTUs, we
expected the MCC scores produced by the OptiFit workflow to be similar
to that of \emph{de novo} clustering using the OptiClust algorithm. In
the OptiFit workflow the test data was fit to the clustered training
data for each of the 100 data splits resulting in an MCC score for each
split of the data. In the remaining workflows, the data was only
clustered once and then split into the training and testing sets
resulting in a single MCC score for each method. Indeed, the MCC scores
were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
OptiFit self-reference workflows (average MCC = 0.879, standard
deviation = 0.002). Previously, we observed that \emph{de novo}
clustering tends to produce higher MCC scores than reference-based
clustering (\protect\hyperlink{ref-sovacool2022}{2}). Consistent with
prior findings the reference-based methods produced lower MCC scores
(OptiFit Greengenes MCC = 0.786; VSEARCH Greengenes MCC = 0.531) than
the \emph{de novo} methods with the same algorithm (OptiClust \emph{de
novo} MCC = 0.884; VSEARCH \emph{de novo} MCC = 0.641). Another metric
we examined for the OptiFit workflow was the fraction of sequences from
the test set that mapped to the reference OTUs. Since sequences that did
not map to reference OTUs were eliminated, if a high percentage of reads
did not map to an OTU we expected this loss of data to negatively impact
classification performance. We found that loss of data was not an issue
since on average 99.8\% (standard deviation = 0.68\%) of sequences in
the test set mapped to the reference OTUs. This number is higher than
the average fraction of reads mapped in the OptiFit Greengenes workflow
( 96.8\% +/- 3.5). These results indicate that the OptiFit
self-reference method performed as well as the OptiClust \emph{de novo}
method and better than using an external database.

We next assessed model performance using OTU relative abundances from
the training data from the workflows to train a model to predict SRNs
and used the model on the held out data. Using the predicted and actual
diagnosis classification, we calculated the area under the receiver
operating characteristic curve (AUROC) for each data split. During
cross-validation (CV) training, the model performance was equivalent
between OptFit self-reference and OptiClust \emph{de novo} (p-value =
0.071; Figure 2A) while while performance for both VSEARCH methods was
lower than the OptiClust \emph{de novo}, OptiFit self, and OptiFit
Greengenes methods (p-values \textless{} 0.05). The trained model was
then applied to the test data classifying samples as either control or
SRN. Both VSEARCH methods perform slightly worse than the OptiClust *de
novo* method (both p-values \textless{} 0.05). However the performance
on the test data was equivalent between the OptiClust \emph{de novo},
OptiFit Greengenes, and OptiFit self-reference approaches (p-value
\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
data could be fit to existing OTU clusters using OptiFit without
impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences from the new samples that do not map to the existing OTU
clusters will be discarded, resulting in a possible loss of information.
However, we demonstrated that OptiFit can be used to fit new sequence
data into existing OTU clusters and it could perform as well in
predicting SRN compared to \emph{de novo} clustering all the sequence
data together. In this instance, the performance of OptiFit was
equivalent to using a database-reference-based method despite the lower
quality of the OTU clusters in the database-reference-based approach.
This likely indicates that the sequences that are important to the model
are well characterized by the reference database. However, a less well
studied system may not be as well characterized by a reference-database
which would make the ability to utilize one's own data a reference an
exciting possiblility. The ability to integrate data from new samples
into existing OTUs enables the implementation of a single machine
learning model. This is important for model implementation because not
all of the data needs to be available or known at the time of model
generation. A robust machine learning model can be implemented as part
of a non-invasive and low-cost diagnostic for SRN and other diseases.

@@

\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
were previously generated from human stool samples. Sequences were
downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
(\protect\hyperlink{ref-baxter2016}{3},
\protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
samples from 490 subjects. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'', while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
and 229 SRN samples.

\textbf{Data processing.} The full dataset was preprocessed with mothur
(v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward and
reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (standard deviation = 4.1) and the test set
an average of 20 times (standard deviation = 4.1`).

\textbf{\emph{Reference-based workflows.}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  OptiFit Self: The preprocess data was split into the training and
  testing sets. The training set was clustered into OTUs using
  OptiClust, then the test set was fit to the OTUs of the training set
  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{2}).
  The OptiFit algorithm was run with method open so that any sequences
  that did not map to the existing OTU clusters would form new OTUs. The
  data was then subsampled to 10,000 reads and any novel OTUs from the
  test set were removed. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\item
  OptiFit Greengenes: Reference sequences from the Greengenes database
  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{9}) were downloaded
  and processed with mothur by trimming to the V4 region and clustered
  \emph{de novo} with OptiClust

  \begin{enumerate}
  \def\labelenumii{\arabic{enumii}.}
  \tightlist
  \item
    The preprocessed data was fit to the clustered reference data using
    OptiFit with the method open to allow any sequences that did not map
    to the existing reference clusters would form new OTUs. The data was
    then subsampled to 10,000 reads and any novel OTUs from the test set
    were removed. The data set was then split into two sets where 80\%
    of the samples were assigned to the training set and 20\% to the
    testing set. This process was repeated for each of the 100 splits
    resulting in 100 training and testing datasets.
  \end{enumerate}
\item
  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
  v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) directly to
  unprocessed Greengenes 97\% OTU reference alignment consistent with
  how VSEARCH is typically used by the QIIME2 software for
  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{5}).
  The data was then subsampled to 10,000 reads and any novel OTUs from
  the test set were removed. The data set was then split into two sets
  where 80\% of the samples were assigned to the training set and 20\%
  to the testing set. This process was repeated for each of the 100
  splits resulting in 100 training and testing datasets.
\end{enumerate}

\textbf{\emph{De novo workflows.}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\item
  OptiClust \emph{de novo}: All the preprocessed data was clustered
  together with OptiClust 1 to generate OTUs. The data was subsampled to
  10,000 reads per sample and the resulting abundance tables were split
  into the training and testing sets. The process was repeated for each
  of the 100 splits resulting in 100 training and testing datasets.
\item
  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) with 97\%
  identity and then subsampled to 10,000 reads per sample. The process
  was repeated for each of the 100 splits resulting in 100 training and
  testing datasets for both workflows.
\end{enumerate}

\textbf{\emph{Machine Learning.}} A random forest model was trained with
the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{10}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale and
center), collapse correlated OTUs, and remove OTUs with zero variance.
The preprocessing from the training set was then applied to the test
set. Any OTUs in the test set that were not in the training set were
removed. P values comparing model performance were calculated as
previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{11}). The
averaged ROC curves were plotted by taking the average and standard
deviation of the sensitivity at each specificity value.

\textbf{\emph{Code Availability.}}

The analysis workflow was implemented in Snakemake
(\protect\hyperlink{ref-koster2012}{12}). Scripts for analysis were
written in R (\protect\hyperlink{ref-R2020}{13}) and GNU bash
(\protect\hyperlink{ref-GNUbash}{14}). The software used includes mothur
v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), VSEARCH v2.15.2
(\protect\hyperlink{ref-rognes2016}{4}), RStudio
(\protect\hyperlink{ref-RStudio2019}{15}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{16}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{17}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{18}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{19}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}

\hypertarget{acknowledgments}{%
\subsection{Acknowledgments}\label{acknowledgments}}

@@

\includegraphics{./figures/Figure1.png}
